---
title: "Installation"
subtitle: "Get arframe running in under two minutes"
execute:
  echo: true
  eval: false
---


## Install arframe

arframe is hosted on GitHub. The fastest way to install it is with `pak`:

```r
# Install pak if you don't have it
install.packages("pak")

# Install arframe from GitHub
pak::pak("vthanik/arframe")
```

Or with `remotes`:

```r
remotes::install_github("vthanik/arframe")
```


## Install pharmaverseadam

The examples in this gallery use the `pharmaverseadam` package, which provides
synthetic CDISC ADaM datasets (CDISCPILOT01 study, 135 subjects):

```r
install.packages("pharmaverseadam")
```

Datasets used in this gallery: `adsl`, `adae`, `adcm`, `advs`, `adlb`, `admh`, `adex`, `adeg`, `adrs_onco`, `adtte_onco`, `adtr_onco`.


## Optional: cards

The **Path 2** (cards) examples use `cards` for ARD-based summarisation:

```r
install.packages("cards")
```


## Optional: Figure Dependencies

The figure examples (Kaplan-Meier, swimmer plot, waterfall plot) use additional packages:

```r
install.packages(c("ggplot2", "survival", "broom", "scales", "patchwork"))
```


## Verify Your Installation

```{r}
#| eval: false
library(arframe)
library(pharmaverseadam)

# Quick sanity check — build and display a small table
adsl_saf <- adsl[adsl$SAFFL == "Y" & adsl$TRT01A != "Screen Failure", ]

adsl_saf[, c("TRT01A", "AGE", "SEX")] |>
  head(10) |>
  fr_table() |>
  fr_titles("Quick Check", "First 10 Safety Population Subjects") |>
  fr_cols(
    TRT01A = fr_col("Treatment", width = 2.5),
    AGE    = fr_col("Age",       width = 0.8, align = "center"),
    SEX    = fr_col("Sex",       width = 0.6, align = "center")
  ) |>
  fr_header(bold = TRUE, align = "center")
```


## Minimal End-to-End Example

This is the smallest complete example that produces a publication-ready table:

```{r}
library(arframe)
library(dplyr, warn.conflicts = FALSE)
library(pharmaverseadam)

adsl_saf <- adsl |>
  filter(SAFFL == "Y", TRT01A != "Screen Failure")

arm_levels <- c("Placebo", "Xanomeline Low Dose", "Xanomeline High Dose")

arm_n <- adsl_saf |>
  filter(TRT01A %in% arm_levels) |>
  count(TRT01A) |>
  pull(n, name = TRT01A)

age_wide <- adsl_saf |>
  filter(TRT01A %in% arm_levels) |>
  group_by(TRT01A) |>
  summarise(
    n          = as.character(n()),
    `Mean (SD)` = sprintf("%.1f (%.2f)", mean(AGE), sd(AGE)),
    Median     = sprintf("%.1f", median(AGE)),
    `Min, Max` = sprintf("%.0f, %.0f", min(AGE), max(AGE)),
    .groups    = "drop"
  ) |>
  tidyr::pivot_longer(-TRT01A, names_to = "Statistic", values_to = "val") |>
  tidyr::pivot_wider(names_from = TRT01A, values_from = val)

age_wide |>
  fr_table() |>
  fr_titles(
    "Table 1",
    "Age Summary",
    "Safety Population"
  ) |>
  fr_cols(
    Statistic = fr_col("Age (years)", width = 2.0),
    !!!setNames(
      lapply(arm_levels, function(a) fr_col(a, align = "decimal")),
      arm_levels
    ),
    .n = arm_n
  ) |>
  fr_header(bold = TRUE, align = "center") |>
  fr_footnotes("Safety Population: all subjects who received at least one dose.")
```


## Render to PDF or RTF

Once you have a `fr_spec` object, render to any format with a single call:

```{r}
# Assign the spec
spec <- age_wide |>
  fr_table() |>
  fr_titles("Table 1", "Age Summary", "Safety Population") |>
  fr_cols(
    Statistic = fr_col("Age (years)", width = 2.0),
    !!!setNames(
      lapply(arm_levels, function(a) fr_col(a, align = "decimal")),
      arm_levels
    ),
    .n = arm_n
  ) |>
  fr_header(bold = TRUE, align = "center")

# Three formats, zero code changes
out_pdf  <- file.path(tempdir(), "age_summary.pdf")
out_rtf  <- file.path(tempdir(), "age_summary.rtf")
out_html <- file.path(tempdir(), "age_summary.html")

spec |> fr_render(out_pdf)
spec |> fr_render(out_rtf)
spec |> fr_render(out_html)

on.exit(unlink(c(out_pdf, out_rtf, out_html)))
```


## Study-Wide Theme

Set a theme once and it applies to every table in the session:

```{r}
fr_theme(
  font_size   = 9,
  font_family = "Courier New",
  orientation = "landscape",
  hlines      = "header",
  n_format    = "{label}\n(N={n})",
  pagehead    = list(left = "Protocol: CDISCPILOT01", right = "CONFIDENTIAL"),
  pagefoot    = list(
    left  = "{program}",
    right = "Page {thepage} of {total_pages}"
  ),
  continuation = "(continued)"
)

# Reset when done
fr_theme_reset()
```


## Rendering to PDF / RTF

Every `fr_spec` object can be rendered to PDF, RTF, or HTML with a single call. No code changes needed between formats:

```{r}
#| eval: false
spec |> fr_render("output.pdf")
spec |> fr_render("output.rtf")
spec |> fr_render("output.html")
```

PDF rendering requires XeLaTeX. See the [arframe documentation](https://github.com/vthanik/arframe) for font setup details.


## Next Steps

- Browse the **Tables** and **Listings** sections in the sidebar for examples.
- Each table page shows two data preparation approaches: manual dplyr and ARD-based cards.
- All examples use the `pharmaverseadam` CDISCPILOT01 synthetic datasets.
