---
title: "Laboratory Results - Chemistry"
subtitle: "Summary Statistics by Visit"
execute:
  echo: true
  eval: true
---


```{r}
#| label: prereqs
#| include: false
library(arframe)
fr_theme(hlines = "header", font_family = "Courier New")
blank_to_na <- function(df) {
  df[] <- lapply(df, function(x) {
    if (is.character(x)) x[x == ""] <- NA_character_
    x
  })
  df
}
```

## Setup

See [Prerequisites](../install.qmd) for installation instructions.

```{r}
#| label: setup
library(arframe)
library(pharmaverseadam)
library(dplyr, warn.conflicts = FALSE)
library(tidyr)
library(cards)

# Chemistry panel — 6 key parameters
param_subset <- c("ALT", "AST", "BILI", "CREAT", "GLUC", "SODIUM")

adlb_saf <- pharmaverseadam::adlb |>
  blank_to_na() |>
  filter(
    SAFFL   == "Y",
    TRT01A  != "Screen Failure",
    PARCAT1 == "CHEMISTRY",
    PARAMCD %in% param_subset,
    ANL01FL == "Y" | AVISIT == "Baseline"
  )

adsl_saf <- pharmaverseadam::adsl |>
  blank_to_na() |>
  filter(SAFFL == "Y", TRT01A != "Screen Failure")

arm_levels <- c("Placebo", "Xanomeline Low Dose", "Xanomeline High Dose")
arm_n      <- adsl_saf |> count(TRT01A) |> pull(n, name = TRT01A)

visit_order <- c(
  "Baseline", "Week 2", "Week 4", "Week 6", "Week 8",
  "Week 12", "Week 16", "Week 20", "Week 24", "Week 26"
)

adlb_saf <- adlb_saf |>
  mutate(
    AVISIT = factor(AVISIT, levels = visit_order),
    TRT01A = factor(TRT01A, levels = arm_levels)
  ) |>
  filter(!is.na(AVISIT))
```


## Data Preparation

::: {.panel-tabset}

### dplyr

```{r}
#| label: dplyr-code

# ── Vertical layout: stats as rows, arms as columns ──
# Per PARAM × AVISIT × TRT01A: N, Mean, SD, Median, Min, Max
lab_stats <- adlb_saf |>
  group_by(PARAM, PARAMCD, AVISIT, TRT01A) |>
  summarise(
    n      = as.character(sum(!is.na(AVAL))),
    Mean   = sprintf("%.1f", mean(AVAL, na.rm = TRUE)),
    SD     = sprintf("%.2f", sd(AVAL, na.rm = TRUE)),
    Median = sprintf("%.1f", median(AVAL, na.rm = TRUE)),
    Min    = sprintf("%.0f", min(AVAL, na.rm = TRUE)),
    Max    = sprintf("%.0f", max(AVAL, na.rm = TRUE)),
    .groups = "drop"
  ) |>
  pivot_longer(
    cols      = c(n, Mean, SD, Median, Min, Max),
    names_to  = "stat_label",
    values_to = "value"
  ) |>
  pivot_wider(names_from = TRT01A, values_from = value) |>
  arrange(PARAM, AVISIT, match(stat_label, c("n", "Mean", "SD", "Median", "Min", "Max"))) |>
  mutate(across(where(is.factor), as.character))

# Ensure all arm columns present
for (arm in arm_levels) {
  if (!arm %in% names(lab_stats)) lab_stats[[arm]] <- ""
}

lab_wide <- lab_stats |>
  select(PARAM, PARAMCD, AVISIT, stat_label, all_of(arm_levels))
```

### cards

This tab showcases `fr_wide_ard()` with **per-variable decimal precision** — different lab parameters get different formatting based on their scale and clinical convention:

```{r}
#| label: cards-code

lab_data <- adlb_saf |>
  mutate(across(where(is.factor), as.character))

lab_ard <- ard_stack(
  data  = lab_data,
  .by   = c("TRT01A", "PARAM", "PARAMCD", "AVISIT"),
  ard_continuous(
    variables = "AVAL",
    statistic = ~ continuous_summary_fns(c("N", "mean", "sd", "median", "min", "max"))
  ),
  .overall = FALSE
) |>
  filter(variable == "AVAL")

# Per-variable decimals: different precision per lab parameter
lab_cards <- fr_wide_ard(
  lab_ard,
  statistic = list(
    continuous = c(
      "n"      = "{N}",
      "Mean"   = "{mean}",
      "SD"     = "{sd}",
      "Median" = "{median}",
      "Min"    = "{min}",
      "Max"    = "{max}"
    )
  ),
  decimals = list(
    ALT    = c(mean = 1, sd = 2, median = 1, min = 0, max = 0),
    AST    = c(mean = 1, sd = 2, median = 1, min = 0, max = 0),
    BILI   = c(mean = 2, sd = 3, median = 2, min = 1, max = 1),
    CREAT  = c(mean = 1, sd = 2, median = 1, min = 0, max = 0),
    GLUC   = c(mean = 1, sd = 2, median = 1, min = 0, max = 0),
    SODIUM = c(mean = 1, sd = 2, median = 1, min = 0, max = 0)
  ),
  column = "TRT01A",
  label  = c(AVAL = "Value")
) |>
  mutate(across(where(is.factor), as.character))
```

:::


## arframe Pipeline

The rendered table below uses the **dplyr** data prep (`lab_wide`). The cards tab produces an equivalent `lab_cards` — swap it in to use the cards path instead.

Stats as rows with treatment arms as columns, grouped by visit, one page per parameter:

```{r}
#| label: pipeline
#| eval: false
lab_wide |>
  fr_table() |>
  fr_titles(
    "Table 14.3.4",
    "Laboratory Results Summary - Chemistry Panel",
    "Safety Population"
  ) |>
  fr_cols(
    PARAMCD    = fr_col(visible = FALSE),
    PARAM      = fr_col(visible = FALSE),
    AVISIT     = fr_col(visible = FALSE),
    stat_label = fr_col("", width = 1.5),
    !!!setNames(
      lapply(arm_levels, function(a) fr_col(a, align = "decimal")),
      arm_levels
    ),
    .n = arm_n
  ) |>
  fr_header(bold = TRUE, align = "center") |>
  fr_rows(
    group_by    = list(cols = "AVISIT", label = "stat_label"),
    blank_after = "AVISIT",
    page_by     = "PARAM",
    group_style = list(bold = TRUE)
  ) |>
  fr_page() |>
  fr_footnotes(
    "N is the number of subjects in the Safety Population per treatment group.",
    "Statistics: n = number of non-missing observations, SD = Standard Deviation.",
    "Chemistry panel: ALT, AST, Bilirubin, Creatinine, Glucose, Sodium."
  )
```


## Rendered Table

```{r}
#| label: rendered-table
#| echo: false
#| ref.label: pipeline
```
