---
title: "Vital Signs"
subtitle: "Values and Change from Baseline by Visit"
execute:
  echo: true
  eval: true
---


```{r}
#| label: prereqs
#| include: false
library(arframe)
fr_theme(hlines = "header", font_family = "Courier New")
blank_to_na <- function(df) {
  df[] <- lapply(df, function(x) {
    if (is.character(x)) x[x == ""] <- NA_character_
    x
  })
  df
}
```

## Setup

See [Prerequisites](../install.qmd) for installation instructions.

```{r}
#| label: setup
library(arframe)
library(pharmaverseadam)
library(dplyr, warn.conflicts = FALSE)
library(tidyr)
library(cards)

# Safety population vitals (baseline + post-baseline analysis visits)
advs_saf <- pharmaverseadam::advs |>
  blank_to_na() |>
  filter(
    SAFFL   == "Y",
    TRT01A  != "Screen Failure",
    PARAMCD %in% c("SYSBP", "DIABP", "PULSE", "TEMP"),
    ANL01FL == "Y" | AVISIT == "Baseline"
  )

# Arm N from ADSL
adsl_saf <- pharmaverseadam::adsl |>
  blank_to_na() |>
  filter(SAFFL == "Y", TRT01A != "Screen Failure")

arm_n      <- adsl_saf |> count(TRT01A) |> pull(n, name = TRT01A)
arm_levels <- c("Placebo", "Xanomeline Low Dose", "Xanomeline High Dose")

# Visit ordering (include Baseline)
visit_order <- c(
  "Baseline", "Week 2", "Week 4", "Week 6", "Week 8",
  "Week 12", "Week 16", "Week 20", "Week 24",
  "Week 26", "End of Treatment"
)

advs_saf <- advs_saf |>
  mutate(
    AVISIT = factor(AVISIT, levels = visit_order),
    TRT01A = factor(TRT01A, levels = arm_levels)
  ) |>
  filter(!is.na(AVISIT))
```


## Data Preparation

::: {.panel-tabset}

### dplyr

```{r}
#| label: dplyr-code

# ── Wide layout: one row per Visit × Treatment ──
# Value stats: N, Mean, SD, Median, Min, Max
# Change from Baseline stats: N, Mean, SD, Median, Min, Max

fmt1 <- function(x) ifelse(is.na(x), "", sprintf("%.1f", x))
fmt0 <- function(x) ifelse(is.na(x), "", sprintf("%.0f", x))

vs_wide <- advs_saf |>
  group_by(PARAM, PARAMCD, AVISIT, TRT01A) |>
  summarise(
    val_n      = as.character(sum(!is.na(AVAL))),
    val_mean   = fmt1(mean(AVAL, na.rm = TRUE)),
    val_sd     = fmt1(sd(AVAL, na.rm = TRUE)),
    val_median = fmt1(median(AVAL, na.rm = TRUE)),
    val_min    = fmt0(min(AVAL, na.rm = TRUE)),
    val_max    = fmt0(max(AVAL, na.rm = TRUE)),
    chg_n      = if (all(is.na(CHG))) "" else as.character(sum(!is.na(CHG))),
    chg_mean   = if (all(is.na(CHG))) "" else fmt1(mean(CHG, na.rm = TRUE)),
    chg_sd     = if (all(is.na(CHG))) "" else fmt1(sd(CHG, na.rm = TRUE)),
    chg_median = if (all(is.na(CHG))) "" else fmt1(median(CHG, na.rm = TRUE)),
    chg_min    = if (all(is.na(CHG))) "" else fmt0(min(CHG, na.rm = TRUE)),
    chg_max    = if (all(is.na(CHG))) "" else fmt0(max(CHG, na.rm = TRUE)),
    .groups    = "drop"
  ) |>
  # Handle visits with no CHG data (Baseline)
 mutate(
    across(starts_with("chg_"), ~ ifelse(.x %in% c("NaN", "NA", "Inf", "-Inf"), "", .x))
  ) |>
  arrange(PARAM, AVISIT, TRT01A) |>
  mutate(across(where(is.factor), as.character))
```

### cards

```{r}
#| label: cards-code

vs_data <- advs_saf |>
  mutate(across(where(is.factor), as.character))

vs_ard <- ard_stack(
  data  = vs_data,
  .by   = c("TRT01A", "PARAM", "AVISIT"),
  ard_continuous(
    variables = c("AVAL", "CHG"),
    statistic = ~ continuous_summary_fns(c("N", "mean", "sd", "median", "min", "max"))
  ),
  .overall = FALSE
) |>
  filter(variable %in% c("AVAL", "CHG"))

# Pivot to wide: one row per Visit × Treatment
vs_cards <- vs_ard |>
  mutate(
    stat_val = as.numeric(unlist(stat)),
    stat_name = as.character(stat_name),
    var = as.character(variable),
    arm = as.character(unlist(group1_level)),
    param = as.character(unlist(group2_level)),
    visit = as.character(unlist(group3_level))
  ) |>
  mutate(
    col_name = paste0(
      ifelse(var == "AVAL", "val_", "chg_"),
      stat_name
    )
  ) |>
  select(param, visit, arm, col_name, stat_val) |>
  pivot_wider(names_from = col_name, values_from = stat_val) |>
  mutate(across(where(is.numeric), ~ ifelse(is.na(.) | is.nan(.) | is.infinite(.), NA_real_, .)))
```

:::


## arframe Pipeline

The rendered table below uses the **dplyr** data prep (`vs_wide`). The cards tab produces an equivalent `vs_cards` — swap it in to use the cards path instead.

Each row is a treatment group at a visit, with Value and Change from Baseline statistics as columns:

```{r}
#| label: pipeline
#| eval: false
vs_wide |>
  fr_table() |>
  fr_titles(
    "Table 14.3.9",
    "Summary of Vital Signs: Values and Change from Baseline",
    "Safety Population"
  ) |>
  fr_cols(
    PARAMCD    = fr_col(visible = FALSE),
    PARAM      = fr_col(visible = FALSE),
    AVISIT     = fr_col(visible = FALSE),
    TRT01A     = fr_col("", width = 1.5),
    val_n      = fr_col("N",      align = "decimal", width = 0.5),
    val_mean   = fr_col("Mean",   align = "decimal", width = 0.7),
    val_sd     = fr_col("SD",     align = "decimal", width = 0.7),
    val_median = fr_col("Median", align = "decimal", width = 0.7),
    val_min    = fr_col("Min",    align = "decimal", width = 0.5),
    val_max    = fr_col("Max",    align = "decimal", width = 0.5),
    chg_n      = fr_col("N",      align = "decimal", width = 0.5),
    chg_mean   = fr_col("Mean",   align = "decimal", width = 0.7),
    chg_sd     = fr_col("SD",     align = "decimal", width = 0.7),
    chg_median = fr_col("Median", align = "decimal", width = 0.7),
    chg_min    = fr_col("Min",    align = "decimal", width = 0.5),
    chg_max    = fr_col("Max",    align = "decimal", width = 0.5)
  ) |>
  fr_spans(
    "Value" = c("val_n", "val_mean", "val_sd", "val_median", "val_min", "val_max"),
    "Change from Baseline" = c("chg_n", "chg_mean", "chg_sd", "chg_median", "chg_min", "chg_max"),
    .gap = FALSE
  ) |>
  fr_header(bold = TRUE, align = "center") |>
  fr_rows(
    group_by    = list(cols = "AVISIT", label = "TRT01A"),
    blank_after = "AVISIT",
    page_by     = "PARAM",
    group_style = list(bold = TRUE)
  ) |>
  fr_page() |>
  fr_footnotes(
    "Value = observed value at visit; Change = post-baseline value minus baseline.",
    "SD = Standard Deviation.",
    "CDISCPILOT01 Safety Population."
  )
```


## Rendered Table

```{r}
#| label: rendered-table
#| echo: false
#| ref.label: pipeline
```
